A Graph Theory Based Approach to MDAO Problem Formulation

### 2013-04-22 graph_syntax.tex
adding in brians updates to section 6

\section{Graph Representation of MDAO Constructs}
\label{s:graph representation}
In this section we will make use of the Sellar problem, described in 
Eqn. \ref{eqn:simple_fpf}. The graph representation of the Sellar problem, 
using the new stynax, is presented in Fig. \ref{f:sellar_graph_full}. Throughout
the rest of this section, different sub-graphs or slight modifications of 
the full graph will be used to demonstrate the important MDAO constructs within
this graph syntax.

@@

\subsection{Analysis Blocks and Connections}
\label{ss:analysis blocks and connections}
Analysis tools take in a number input variables and then perform some work to calculate 
the values for their respective outputs. In an MDAO graph, this process is 
represented by a group of nodes and edges called an \emph{analysis block}, 
shown in Fig. \ref{f:analysis block}. Within an analysis block each variable 
node represents a single input or output and is connected 
to a single model node via a fixed edge. Note that in Fig.~\ref{f:analysis block}, 
the analysis block contains two model nodes, with a single edge connecting them. 
This edge represents the necessary calculations to map given inputs 
to proper output values. If constructing a weighted graph, the weight calculation edge 
provides the opportunity to computational cost. However, it is also allowed to 
skip this edge and have both incomming and outgoing edges to variables through a 
single model node for a given analysis block. Although analysis blocks are 
comprised of multiple nodes and edges, since all the edges within them are 
fixed, they become fixed sub-graphs within a graph. In this way, they cannot be 
altered but they can be removed as a group.

@@

Variable nodes in an analysis block can be distinguished as either an input or 
an output by the manner in which they are connected. As shown in Fig.~\ref{f:analysis block}, 
inputs are any variable nodes that have an outgoing edge into a model 
node. Conversely, outputs are any variable nodes that have an incoming edge from a model node.

MDAO problems require that information be passed between sets of analyses. When 
information from the output of one analysis block is passed, or connected, to the 
input of another analysis block, a new connection edge is added connecting the two 
variable nodes involved in the exchange. These connection edges are free, unlike the edges 
within an analysis block, can be added or removed depending on the needs
of a given design problem. In other words, free connection edges are created by 
engineers linking the output of one tool to the input of another one. Note that 
it is allowed for a single output to have outgoing connection edges to multiple 
downstream inputs.

\subsection{Design Variables}
For the Sellar problem, there are three design variables, $x_1^*$, $z_1^*$, and $z_2^*$. In Fig. 
\ref{f:sellar_graph_full}, these are the only three holes (nodes without out 
incomming edges) in the graph.

\begin{figure}[htb!]
  \begin{center}
  [NOTE: Re-work this one to match $D_2$ from Sellar]
    \includegraphics[width=.6\textwidth]{images/design_vars_graph}
  \end{center}
  \caption{Notional graph with two potential design variables \label{f:designvars}}
\end{figure}

More formally stated, given a MCG with the $\txt{deg}_l^-(v)=1$ for all nodes, 
there will likely be a set of holes preventing a valid FPG from being obtained. 
Each hole represents a node that could potentially become a design variable. 
It is up to the designer to examine each and decide if it is appropriate to 
consider it a design variable, in which case the designer shall set the lower 
indegree limit to be zero. However, some variable holes may not actually be 
suitable as a design variable. For instance, if an aircraft mission analysis 
code requires a drag input which ends up being a hole then it's likely best to 
leave$\txt{deg}_l^-(v)=1$ and find a new analysis code to fill the hole in graph.

So a design variable is defined as any variable node with $\txt{deg}_l^-(v)=0$. 
This, by definition, means that design variables are not holes in the graph, since 
they have not violated their lower indegree limit. Regardless, 
when working with a MCG or FPG, only connection edges are allowed, so a design 
variable in either graph will be a variable with no incomming edges edges. For a valid PSG 
any design variable node would have at least one, but possibly multiple, incomming 
driven edges from one or more driver nodes.

\subsection{Objectives, and Constraints}
\label{ss:objectives and constraints}
Like design variables, objectives and constraints are also constructs that need 
to be identified when setting up a given design problem. In the case of 
objective functions single output values could be selected, but commonly multiple 
values are assembeled together via simple algebraic expressions to form a composite 
objective function. (e.g. The objective function from Eqn. \ref{eqn:simple_fpf} is
$x_1^2+z_2+y_1+e^{-y_2}$). Constraints are always given in the 
form of either an inequality or an equality. (e.g. a constraint from Eqn. 
\ref{eqn:simple_fpf} is $\frac{y_1}{3.16}-1\geq0$.) By convention, 
constraints are usually given such that some expression will be less than or equal 
to zero, so any positive value would violate the constraint. For both objective 
and constraints, these simple expressions take some inputs and map them to an
output value of significance to the overall design problem.

These operations, though very simple, are effectively the same kind of task 
performed by an analysis block. A constraint or objective function can therfore 
be represented as another analysis block on the graph, with it's own input and 
output variable nodes. Although fundamentally no different than an analysis block, 
for clarity and convience it is useful to distinguish between regular analysis 
blocks and those that arise from the addition of objectives or constraints to 
the graph. Therefore, we define an \emph{expression block}, as the collection of variable and model 
nodes related to a given objective or constraint function. Fig. \ref{f:obj-cons}
highlights the objective and constraint blocks from the Sellar problem.

@@

  A simple definition of a global node is any node that involves information 
  from more than one analysis block. Likewise, a local node is any node 
  that involves information from only a single analysis block. As stated before, variable 
  nodes with fixed edges directed to model nodes (as part of an analysis block) are inherently local 
  and similarly model nodes are also local. Global variables are a part of many 
  MDAO problems and must be representable in the graph. Follwoing our simple definition
  any variable node that had multiple incomming or outgoing edges, would be global. 
  In Fig. \ref{f:sellar_graph_full} the variable nodes for $x_1^*$, $z_1^*$, and $z_2^*$ 
  have multiple outgoing edges and would be considered global.

  Note that the variable $x_1$ from the Sellar problem is usually refered to as 
  a local variable, which is contradicted by the given graph. This discrepancy 
  arises from the explicit treatment of objectives and constraints as separate 
  expression blocks. Normally in the Sellar problem $x_1$ is considered a local 
  variable because it only directly affects analysis block $D_1$. By forcing
  the expansion of $F$ into an expression block with its own inputs, then $F$ 
  gets it's own copy of the $x_1$ variable. This necessitates the creation 
  of the global $x_1^*$ node in the graph with two outgoing edges to link the 
  two $x_1$ inputs nodes in the different blocks. Interestingly, using this 
  definition, where $x_1^*$ is a global variable fits nicely within 
  the structure of Braun's Colaborative Optimization (CO) archtiecture \cite{braun1996thesis}. 
  The rules of setting up a problem with CO require that each discipline be 
  given an independent local copy of all global variables. All local 
  variables are retained uniquely within their respective disciplines, except 
  when a local design variable shows up explicitely in the objective 
  or global constraint. In that case a global variable is created, and the discipline again 
  gets an independent local copy of it. Given our definition of global vs local
  variable nodes, no such exception for local variables in global expressions is 
  necessary. When you include a given variable in an expression block, it becomes 
  global by definition and should be treated as such in CO.

  Analysis and expression blocks can also have a locality defined for the set of 
  nodes and edges within the block. Their locality is determined by the 
  incomming and outgoing edges from the block as a whole. Fig. 
  \ref{f:sellar_graph_full} has expression blocks for each constraint, 
  $G_1$ and $G_2$. Both constraints are local to their respective analysis 
  blocks. But the expression block for $F$ has incomming edges from $D_1$ and 
  $D_2$, so it would considered global.

>>>>

\section{Graph Representation of MDAO Constructs}
\label{s:graph representation}
In this section we present our graph syntax applied to the description of MDAO 
Problems. We make use of the Sellar problem, described in 
Eqn. \ref{eqn:simple_fpf} as an example to illustrate the implementation of the syntax. 
The graph representation of the Sellar problem, using the new stynax, is 
presented in Fig. \ref{f:sellar_graph_full}. In this section sub-graphs or 
slight modifications of the full graph will be used to demonstrate the 
important MDAO constructs described by this graph syntax.

@@

\subsection{Analysis Blocks and Connections}
\label{ss:analysis blocks and connections}
Analysis tools take in a number input variables and then calculate 
the values for their respective outputs. We represent this process
by a group of nodes and edges called an \emph{analysis block}, 
shown in Fig. \ref{f:analysis block}. Within an analysis block, each variable 
node represents a single input or output and is connected 
to a single model node via a fixed edge. Note that in Fig.~\ref{f:analysis block}, 
the analysis block includes two model nodes, with a single edge connecting them. 
This edge represents the necessary calculations to map given inputs 
to the outputs of the anlysis. This calculation edge provides the opportunity 
to encode computational cost or other characteristics of the analysis code as 
a weight in a weighted graph formulation. If associating a weight is 
unnecessary, the calculation edge may be omitted and all input and output 
variables can be connected to a single model node for a given analysis block. 
Since all the edges within analysis blocks are fixed, the blocks themselves are 
fixed sub-graphs within the overall MDAO problem graph. The connectivity of 
nodes and edges in an analysis block cannot be altered for the purposes of MDAO 
problem formulation; however, analysis block sub-graphs can be added or removed 
from the MDAO problem graph as needed.

@@

Variable nodes in an analysis block can be distinguished as either an input or 
an output by the manner in which they are connected. As shown in Fig.~\ref{f:analysis block}, 
inputs are represented as variable nodes that have an outgoing edge into a model 
node. Conversely, outputs represented as any variable nodes that have an incoming 
edge from a model node.

MDAO problems require that information be passed between sets of analyses. When 
information from one output of an analysis block is passed, or connected, to an 
input of another analysis block, a new connection edge is added between the two 
variable nodes involved in the exchange. These connection edges are free, and unlike the edges 
within an analysis block, the edges can be added or removed depending on the needs
of a given design problem. In other words, free connection edges are created by 
engineers linking an output of one tool to an input of another one. It is 
allowable for a single output to have outgoing connection edges to multiple 
downstream inputs.

\subsection{Design Variables}
Design variables are the input variables that are free to be 
changed by the designer or by an optimizer in an MDAO study. In a graph design
variables will be a set of variable nodes, $V_{dv}$, identified as follows:

\begin{equation}
  V_{dv} = \{V \st \txt{deg}_l^-(v)=0\}
\end{equation}

For the Sellar problem, there are three design variables: $x_1^*$, $z_1^*$, and 
$z_2^*$. In Fig. \ref{f:designvars} all variable from the Sellar problem are 
annotated with their lower indegree limit and the three design variables are identified 
in bold.

\begin{figure}[htb!]
  \begin{center}
    [Graph Here....]
    \includegraphics[width=.6\textwidth]{images/design_vars_graph}
  \end{center}
  \caption{Sellar problem graph with lower indegree limits for all variables \label{f:designvars}}
\end{figure}

In Fig. \ref{f:designvars}, none of the design variable nodes have any incomming 
edges. In general, in a MCG or a FPG, design variables will 
show up as nodes without any incomming edges. For a valid PSG, design variables 
could have one or more incomming driven edges representing a value determined by 
a optimizer.

Note the although there are no incomming edges, design variables are not holes 
in the graph as defined by Eq. \ref{e:hole}. Since they have $\txt{deg}_l^-(v)=0$ 
having no incomming connection edges does not violate their lower indegree limit. 
However, the concept of holes can be useful for identifying potential design variables
given an arbitrary MCG. If you first set $\txt{deg}_l^-(v)=1$ for all variable nodes, 
and then find the resulint holes, you will have a list of variable nodes that 
could potentially be design variables. Some variables will be true design variables, 
while others might represent missing information in the problem formulation. For 
instance, if there is a hole for a drag input of an aircraft mission analysis code 
then this is not likely a good choice for a design variable. So it is up to the designer to 
examine each hole and decide if it's an appropriate design variable. If 
the variable should become a design variable then set $\txt{deg}_l^-(v)=0$. 
If a variable should not be a design variable, then leave it as a hole which 
needs to be addressed in a manner discussed in section \ref{ss:obtaining FPG}.

\subsection{Objectives and Constraints}
\label{ss:objectives and constraints}
Objectives and constraints are constructs that need to be identified when 
describing and MDAO problem. In the case of objective functions a single output 
value generated by an analysis block could be selected, but commonly multiple 
values are assembeled together to form a composite objective function. 
Generally both objective and constraint functions accept a set of inputs and map 
them to an output value of significance to the overall design problem.

The operations of implementing composite objective and constraint functions, 
although typically simple, are effectively the identical to the task performed by an 
analysis block. A composit constraint or objective function can therfore 
be represented as an additional analysis block within the graph, with its own input and 
output variable nodes. Although fundamentally no different than an analysis block, 
for clarity and convience, it is useful to distinguish between analysis 
blocks corresponding to analysis codes and those that arise from the addition of 
objectives or constraints to the graph. Therefore, we define an 
\emph{expression block}, as the collection of variable and model nodes related 
to a given objective or constraint function. Fig. \ref{f:obj-cons} highlights 
the objective and constraint blocks from the Sellar problem. If the relevant 
objective and constraint values are direct results from other analysis blocks 
and composite functions are not needed, then expression blocks may be omitted 
from the MDAO problem graph.

@@

  The set of local nodes, $V_l$, is the collection of 
  nodes with exactly one outgoing connection edge.

  \begin{equation}
    V_{l} = \{V \st \txt{deg}^+(v)=1\}
    \label{e:local_definition}
  \end{equation}

  In section \ref{s:specific vs fundamental} variable nodes that are part of 
  an analysis block are inherently local and similarly model nodes 
  are also local. Following the definition in Eq. \ref{e:local_definition}
  we can now see why. Variable nodes connected to model nodes are only allowed to 
  have one outgoing edge, and hence will always by definition be local variables.

  The set of global nodes, $V_g$, is the collection of all nodes with more than 
  one outgoing connection edge.

  \begin{equation}
    V_{g} = \{V \st \txt{deg}^+(v)>1\}
    \label{e:global_definition}
  \end{equation}

  Follwoing our definition any variable node that had multiple 
  outgoing edges, is global. In Fig. \ref{f:sellar_graph_full} 
  the variable nodes for $x_1^*$, $z_1^*$, and $z_2^*$ have multiple outgoing 
  edges and would be considered global.

  Note that the variable $x_1$ from the Sellar problem is usually refered to as 
  a local variable, which is contradicted by the given graph. This discrepancy 
  arises from the explicit treatment of objectives and constraints as separate 
  expression blocks. Normally in the Sellar problem, $x_1$ is considered a local 
  variable because it directly affects only analysis block $D_1$. By expanding
  $F$ into an expression block with its own inputs, then $F$ is assigned its own local 
  copy of the $x_1$ variable. This formulation necessitates the creation of the global $x_1^*$ 
  node in the graph with two outgoing edges to link the two $x_1$ inputs nodes 
  in the different blocks. Interestingly, the classification of $x_1^*$ 
  as a global variable fits nicely within the structure of the Colaborative 
  Optimization (CO) archtiecture \cite{braun1996thesis}. The rules for setting up 
  a problem with CO require that each discipline be given an independent local 
  copy of all global variables. All local variables are retained uniquely 
  within their respective disciplines, except when a local design variable appears 
  explicitely in the objective or global constraint. In that case a global 
  variable is created, and the analysis block again is again assigned an 
  independent local copy of it. Given our definition of global vs local variable 
  nodes, no such exception for local variables in global expressions is 
  necessary. Whenever a given variable is included in an expression block, 
  it forces the creation of a global variable node in the graph which will 
  result in the correct structure for CO automatically.

  Analysis and expression blocks can themselves be local or global, but the 
  criteria for determining this is slightly different than for variable nodes. 
  In this case, you consider the set of connection edges for the block
  (e.g. from all the variable nodes within the block) as a set. Additionally, 
  you must consider both incoming and outgoing edges, where for variables 
  only outgoing connection edges were relevant.

  A local analysis block is defined as a block with incomming edges from 
  exactly one other analysis block and outgoing edges from exactly one other analysis 
  block.

  A global analysis block is defined as

  Fig. 
  \ref{f:sellar_graph_full} has expression blocks for each constraint, 
  $G_1$ and $G_2$. Both constraints are local to their respective analysis 
  blocks. But the expression block for $F$ has incomming edges from $D_1$ and 
  $D_2$, so it would considered global.
